---
title: "Introduction to `carpenter`"
author: "Luke W. Johnston"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to carpenter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

```{r, eval=FALSE}

# Possible datasets to use:
# head(iris)
# head(airquality)
# head(longley)
# head(swiss)

library(dplyr)
library(tidyr)
ds <- data.frame(state.x77, Region = state.region) %>% 
    mutate(Dangerous = ifelse(Murder > 10, 'Yes', 'No'),
           Poor = ifelse(Income < 4000, 'Yes', 'No'))
ds[c(1, 5, 15, 25, 36), c('Dangerous')] <- NA 
outline_table(ds, c('Population', 'Life.Exp', 'Dangerous', 'Region'), c('Poor')) %>% 
    add_rows(c('Dangerous', 'Region')) %>% 
    add_rows(c('Population', 'Life.Exp')) %>% 
    rename_rows(function(x) gsub('North', 'Nord', x)) %>% 
    rename_columns('Rows', 'Yes', 'NOPE!!') %>% 
    construct_table()

```

